\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in \proglang{R}

### 2015-04-13 papers/jss/JSSemails/response-to-reviewers.tex

\pointRaised{Comment 1}{The paper gives an overview of the RProtoBuf package which implements an 
  R interface to the Protocol Buffers library for an efficient 
  serialization of objects. The paper is well written and easy to read. 
  Introductory code is clear and the package provides objects to play with 
  immediately without the need to jump through hoops, making it appealing. 
  The software implementation is executed well.}
\reply{Thank you.}

\pointRaised{Comment 2}{There are, however, a few inconsistencies in the implementation and some 
  issues with specific sections in the paper. In the following both issues 
  will be addressed sequentially by their occurrence in the paper.}
\reply{Done. These and others have been identified and addressed.  Thank you
  for taking the time to enumerate these issues.}

\pointRaised{Comment 3}{p.4 illustrates the use of messages. The class implements list-like 
  access via \texttt{[[} and \$, but strangely \texttt{names()} return NULL and \texttt{length() }
  doesn't correspond to the number of fields leading to startling results like
the following:}

\begin{verbatim}
 > p
[1] "message of type 'tutorial.Person' with 2 fields set"
 > length(p)
[1] 2
 > p[[3]]
[1] ""
\end{verbatim}
\reply{Done. We have corrected the list-like accessor, fixed \texttt{length()} to
  correspond to the number of set fields, and added \texttt{names()}:}
\begin{verbatim}
> p
message of type 'tutorial.Person' with 0 fields set
> length(p)
[1] 0
> p[[3]]
[1] ""
> p$id <- 1
> length(p)
[1] 1
> names(p)
[1] "name"  "id"    "email" "phone"
\end{verbatim}

@@

\pointRaised{Comment 3 cont.}{The inconsistencies get even more bizarre with descriptors (p.9):}

\begin{verbatim}
 > tutorial.Person$email
[1] "descriptor for field 'email' of type 'tutorial.Person' "
 > tutorial.Person[["email"]]
Error in tutorial.Person[["email"]] : this S4 class is not subsettable
 > names(tutorial.Person)
NULL
 > length(tutorial.Person)
[1] 1
\end{verbatim}
\reply{Done. We agree, and have addressed this inconsistency.  Thank you for
  catching this.}
\begin{verbatim}
> tutorial.Person$email
descriptor for field 'email' of type 'tutorial.Person' 
> tutorial.Person[["email"]]
descriptor for field 'email' of type 'tutorial.Person' 
> names(tutorial.Person)
[1] "name"        "id"          "email"       "phone"       "PhoneNumber"
[6] "PhoneType"  
> length(tutorial.Person)
[1] 6
\end{verbatim}

\pointRaised{Comment 3 cont.}{It appears that there is no way to find out the fields of a descriptor 
  directly (although the low-level object methods seem to be exposed as 
  \texttt{\$field\_count()} and \texttt{\$fields()} - but that seems extremely cumbersome). 
  Again, implementing names() and subsetting may help here.}
\reply{Done. We have implemented \texttt{names} and subsetting.  Thank you for the
  suggestion.}
\begin{verbatim}
> tutorial.Person[[1]]
descriptor for field 'name' of type 'tutorial.Person' 
> tutorial.Person[[2]]
descriptor for field 'id' of type 'tutorial.Person' 
\end{verbatim}

@@

\pointRaised{Comment 4}{Another inconsistency concerns the \texttt{as.list()} method which by design 
  coerces objects to lists (see \texttt{?as.list}), but the implementation for 
  EnumDescriptor breaks that contract and returns a vector instead:}

\begin{verbatim}
 > is.list(as.list(tutorial.Person$PhoneType))
[1] FALSE
 > str(as.list(tutorial.Person$PhoneType))
  Named int [1:3] 0 1 2
  - attr(*, "names")= chr [1:3] "MOBILE" "HOME" "WORK"
\end{verbatim}

\reply{Done, thank you. New output below:}
\begin{verbatim}
> is.list(as.list(tutorial.Person$PhoneType))
[1] TRUE
> str(as.list(tutorial.Person$PhoneType))
List of 3
 $ MOBILE: int 0
 $ HOME  : int 1
 $ WORK  : int 2
\end{verbatim}

\pointRaised{Comment 4 cont}{As with the other interfaces, names() returns NULL so it is again quite
  difficult to perform even simple operations such as finding out the 
  values. It may be natural use some of the standard methods like names(), 
  levels() or similar. As with the previous cases, the lack of [[ support
  makes it impossible to map named enum values to codes and vice-versa.}
\reply{Done, thank you.  New output:}
\begin{verbatim}
> names(tutorial.Person$PhoneType)
[1] "MOBILE" "HOME"   "WORK"  
> tutorial.Person$PhoneType[["HOME"]]
[1] 1
\end{verbatim}

\pointRaised{Comment 5}{In general, the package would benefit from one pass of checks to assess
  the consistency of the API. Since the authors intend direct interaction
  with the objects via basic standard R methods, the classes should behave 
  consistently.}
\reply{We made several passes, correcting issues as documented in the
  \texttt{ChangeLog} and now present in our latest 0.4.2 release on CRAN.}

@@

\pointRaised{Comment 6}{Finally, most classes implement coercion to characters, which is not 
  mentioned and is not quite intuitive for some objects. For example, one
  may think that \texttt{as.character()} on a file descriptor returns let's say the 
  filename, but we get:}

\begin{verbatim}
 > cat(as.character(tutorial.Person$fileDescriptor()))
syntax = "proto2";

package tutorial;

option java_package = "com.example.tutorial";
option java_outer_classname = "AddressBookProtos";
[...]
\end{verbatim}
\reply{The behavior is documented in the package documentation but
  seemed like a minor detail not important for an already-long paper.
  In choosing the debug output for a file descriptor we agree
  that \texttt{filename} is a reasonable thing to expect, but we also
  think that the contents of the \texttt{.proto} file is also
  reasonable, but more useful.  We document this in the help for
  ``FileDescriptor-class'', the vignette, and other sources.
  \texttt{@filename} is one of the slots of the FileDescriptor class
  and so very easy to find.  The contents of the \texttt{.proto} are
  not as easily accessible in a slot, however, and so we find it much
  more useful to be output with \texttt{as.character()}.}

\pointRaised{Comment 7}{It is not necessary clear what java\_package has to do with a file 
  descriptor in R. Depending on the intention here, it may be useful to 
  explain this feature.
}
\reply{Done. This snippet has been removed as part of the general move of
  less relevant details to the package documentation, but for
  reference the \texttt{.proto} file syntax is defined in the Protocol Buffers
  language guide which is referenced earlier. It is a cross platform
  library and so this syntax specifies some parameters when Java code
  is used to access the structures defined in this file.  No such
  special syntax is required in the \texttt{.proto} files for R
  language code and so this line about java\_package was not relevant
  or needed in any way for RProtoBuf and is documented elsewhere.}

@@

\subsubsection*{Other comments:}

\pointRaised{Comment 8}{p.17: "does not support ... function, language or environment. Such 
  objects have no native equivalent type in Protocol Buffers, and have 
  little meaning outside the context or R"
  That is certainly false. Native mirror of environments are hash tables - 
  a very useful type indeed. Language objects are just lists, so there is
  no reason to not include them - they can be useful to store expressions
  that may not be necessary specific to R. Further on p. 18 your run into
  the same problem that could be fixed so easily.}
\reply{Acknowledged.  Environments are more than just hash
  tables because they include other configuration parameters that are
  necessary to serialize as well to make sure
  serialization/unserialization is indempotent, but we agree it is
  cleaner and the package and the exposition in the paper to just make
  sure we serialize everything.  We can now fall back to
  \texttt{base::serialize()} and storing the bits in a rawString type of
  RProtoBuf to make the R schema-less serialization more complete.}

@@

\pointRaised{Comment 9}{The examples in sections 7 and 8 are somewhat weak. It does not seem 
  clear why one would wish to unleash the power of PB just to transfer 
  breaks and counts for plotting - even a simple ASCII file would do that
  just fine. The main point in the example is presumably that there are 
  code generation methods for Hadoop based on PB IDL such that Hadoop can
  be made aware of the data types, thus making a histogram a proper record 
  that won't be split, can be combined etc. -- yet that is not mentioned 
  nor a way presented how that can be leveraged in practice. The Python 
  example code simply uses a static example with constants to simulate the 
  output of a reducer so it doesn't illustrate the point - the reader is 
  left confused why something as trivial would require PB while a savvy 
  reader is not able to replicate the illustrated process. Possibly 
  explaining the benefits and providing more details on how one would 
  write such a job would make it much more relevant.}
\reply{Yes, we added more detail about the advantages of using a
  proper data type for the histograms in this example that you mentioned here -- the
  ability to write combiners, prevent arbitrary splitting of the
  records, etc that can greatly improve performance.  We agree with
  the other reviewer that we don't want to get bogged down in details
  about a particular MapReduce implementation (such as Hadoop) and so
  now we specifically mention that goal here.
  I think we make a better connection now between the
  abstract MapReduce example given, and then the simpler Python
  example code with a static example.}

\pointRaised{Comment 10}{Section 8 is not very well motivated. It is much easier to use other 
  formats for HTTP exchange - JSON is probably the most popular, but even
  CSV works in simple settings. PB is a much less common standard. The 
  main advantage of PB is the performance over the alternatives, but HTTP
  services are not necessarily known for their high-throughput so why one
  would sacrifice interoperability by using PB (they are still more hassle 
  and require special installations)? It would be useful if the reason 
  could be made explicit here or a better example chosen.}
\reply{Done. This section has been reworded to make it shorter and more
  crisp, with fewer extraneous details about OpenCPU. Protocol
  Buffers is an efficient protocol used between distributed systems at
  many of the world's largest internet companies (Twitter, Sony,
  Google, etc.) but the design and implementation of a large
  enterprise-scale distributed system with a complex RPC system and
  serialization needs is well beyond the scope of what we can add to a
  paper about RProtoBuf.  We chose this example because it is a much
  more accessible example that any reader can use to easily
  send/receive RPCs and parse the results with RProtoBuf.}

### 2016-07-10 papers/jss-finalising/v71i02.orig.tex
finalising: content of emailed zip file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@952 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[article]{jss}
\usepackage{booktabs}
\usepackage{listings}
\usepackage[toc,page]{appendix}

% Line numbers for drafts.
%\usepackage[switch, modulo]{lineno}
%\linenumbers

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Spelling Standardization:
% Protocol Buffers, not protocol buffers
% large-scale, not large scale
% Oxford comma: foo, bar, and baz.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%
% Local helpers to make this more compatible with R Journal style.
%
\RequirePackage{fancyvrb}
\RequirePackage{alltt}
\DefineVerbatimEnvironment{example}{Verbatim}{}
% Articles with many authors we should shorten to FirstAuthor, et al.
\shortcites{sciencecloud,janus,dremel,nlme}
\author{Dirk Eddelbuettel\\Debian Project \And 
        Murray Stokely\\Google, Inc \And
        Jeroen Ooms\\UCLA}
\title{\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Dirk Eddelbuettel, Murray Stokely, Jeroen Ooms} %% comma-separated
\Plaintitle{RProtoBuf: Efficient Cross-Language Data Serialization in R}
\Shorttitle{\pkg{RProtoBuf}: Protocol Buffers in \proglang{R}} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
  Modern data collection and analysis pipelines often involve
  a sophisticated mix of applications written in general purpose and
  specialized programming languages.  
  Many formats commonly used to import and export data between
  different programs or systems, such as \code{CSV} or \code{JSON}, are
  verbose, inefficient, not type-safe, or tied to a specific programming language.
  Protocol Buffers are a popular
  method of serializing structured data between applications---while remaining
  independent of programming languages or operating systems.
  They offer a unique combination of features, performance, and maturity that seems
  particularly well suited for data-driven applications and numerical
  computing.
  The \pkg{RProtoBuf} package provides a complete interface to Protocol
  Buffers from the
  \proglang{R} environment for statistical computing.
  This paper outlines the general class of data serialization
  requirements for statistical computing, describes the implementation
  of the \pkg{RProtoBuf} package, and illustrates its use with
  example applications in large-scale data collection pipelines and web
  services.
  %% TODO(ms) keep it less than 150 words. -- I think this may be 154,
  %% depending how emacs is counting.
}
\Keywords{\proglang{R}, \pkg{Rcpp}, Protocol Buffers, serialization, cross-platform}
\Plainkeywords{R, Rcpp, Protocol Buffers, serialization, cross-platform} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Dirk Eddelbuettel \\
  Debian Project \\
  River Forest, IL, USA\\
  E-mail: \email{edd@debian.org}\\
  URL: \url{http://dirk.eddelbuettel.com}\\
  \\
  Murray Stokely\\
  Google, Inc.\\
  1600 Amphitheatre Parkway\\
  Mountain View, CA, USA\\
  E-mail: \email{mstokely@google.com}\\
  URL: \url{http://www.stokely.org/}\\
  \\
  Jeroen Ooms\\
  UCLA Department of Statistics\\
  University of California\\
  Los Angeles, CA, USA\\
  E-mail: \email{jeroen.ooms@stat.ucla.edu}\\
  URL: \url{https://jeroenooms.github.io}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\SweaveOpts{concordance=FALSE,prefix.string=figures/fig}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

% We don't want a left margin for Sinput or Soutput for our table 1.
%\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=0em}
%\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=0em}
%\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}
% Setting the topsep to 0 reduces spacing from input to output and
% improves table 1.
\fvset{listparameters={\setlength{\topsep}{0pt}}}
\renewenvironment{Schunk}{\vspace{\topsep}}{\vspace{\topsep}}

%% DE: I tend to have wider option(width=...) so this
%%     guarantees better line breaks
<<echo=FALSE,print=FALSE>>=
## cf http://www.jstatsoft.org/style#q12
options(prompt = "R> ", 
        continue = "+  ", 
        width = 70, 
        useFancyQuotes = FALSE, 
        digits = 4)
@

\maketitle

@@

Modern data collection and analysis pipelines increasingly involve collections
of decoupled components in order to better manage software complexity 
through reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
These pipelines are frequently built using different programming 
languages for the different phases of data analysis --- collection,
cleaning, modeling, analysis, post-processing, and
presentation --- in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may produce intermediate results that need to be
stored in a file, or sent over the network for further processing.

Given these requirements, how do we safely and efficiently share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer systems?
In computer programming, \emph{serialization} is the process of
translating data structures, variables, and session state into a
format that can be stored or transmitted and then reconstructed in the
original form later \citep{clinec++}.
Programming
languages such as \proglang{R}, \proglang{Julia}, \proglang{Java}, and \proglang{Python} include built-in
support for serialization, but the default formats 
are usually language-specific and thereby lock the user into a single
environment.

Data analysts and researchers often use character-separated text formats such
as \code{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \code{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as separator or decimal point vary by
country.  \emph{Extensible Markup Language} (\code{XML}) is a
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexity at the parsing side (these problems are somewhat mitigated by the
availability of mature libraries and parsers). Because \code{XML} is
text-based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric data sets as they appear in statistical
applications.

A more modern format is \emph{JavaScript ObjectNotation}
(\code{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and already widely-used on the world wide web.
Several \proglang{R} packages implement functions to parse and generate
\code{JSON} data from \proglang{R} objects \citep{rjson,RJSONIO,jsonlite}.
\code{JSON} natively supports arrays and four primitive types: numbers, strings,
booleans, and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is inefficient and
leads to loss of type (double versus integer) and precision. 
A number of binary formats based on \code{JSON} have been proposed
that reduce the parsing cost and improve efficiency, but these formats
are not widely supported.  Furthermore, such formats lack a separate
schema for the serialized data and thus still duplicate field names
with each message sent over the network or stored in a file.

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift \citep{Apache:Thrift}, and Apache Avro \citep{Apache:Avro}
provide a compact well-documented schema for cross-language data
structures and efficient binary interchange formats.  Since the schema
is provided separately from the data, the data can be
efficiently encoded to minimize storage costs when
compared with simple ``schema-less'' binary interchange formats.
%Many sources compare data serialization formats
%and show Protocol Buffers perform favorably to the alternatives; see
%\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.
Protocol Buffers performs well in the comparison of such formats by
\citet{Sumaray:2012:CDS:2184751.2184810}.

This paper describes an \proglang{R} interface to Protocol Buffers,
and is organized as follows. Section~\ref{sec:protobuf}
provides a general high-level overview of Protocol Buffers as well as a basic
motivation for their use.
Section~\ref{sec:rprotobuf-basic} describes the interactive \proglang{R} interface
provided by the \pkg{RProtoBuf} package, and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
details the implementation of the main S4 classes and methods.
Section~\ref{sec:types} describes the challenges of type coercion
between \proglang{R} and other languages.  Section~\ref{sec:evaluation} introduces a
general \proglang{R} language schema for serializing arbitrary \proglang{R} objects and compares it to
the serialization capabilities built directly into \proglang{R}.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \pkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.

@@

Protocol Buffers are a modern, language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data. Some of their
features, particularly in the context of data analysis, are:

\begin{itemize}
\item \emph{Portable}:  Enable users to send and receive data between
  applications as well as different computers or operating systems.
\item \emph{Efficient}:  Data is serialized into a compact binary
  representation for transmission or storage.
\item \emph{Extensible}:  New fields can be added to Protocol Buffer schemas
  in a forward-compatible way that does not break older applications.
\item \emph{Stable}:  Protocol Buffers have been in wide use for over a
  decade.
\end{itemize}

%\begin{figure}[bp]
\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/protobuf-distributed-system-crop.pdf}
\end{center}
\caption{Example usage of Protocol Buffers.}
\label{fig:protobuf-distributed-usecase}
\end{figure}

Figure~\ref{fig:protobuf-distributed-usecase} illustrates an example
communication work flow with Protocol Buffers and an interactive \proglang{R} session.
Common use cases include populating a request remote-procedure call (RPC)
Protocol Buffer in \proglang{R} that is then serialized and sent over the network to a
remote server.  The server deserializes the message, acts on the
request, and responds with a new Protocol Buffer over the network.
The key difference to, say, a request to an \pkg{Rserve}
\citep{Urbanek:2003:Rserve,CRAN:Rserve} instance is that
the remote server may be implemented in any language.
%, with no dependence on \proglang{R}.

While traditional IDLs have at times been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is flexible and easy to use.  The schema for structured
Protocol Buffer data is defined in \code{.proto} files, which may
contain one or more message types.  Each message type has one or more
fields.  A field is specified with a unique number (called a \emph{tag number}), a name, a value
type, and a field rule specifying whether the field is optional,
required, or repeated.  The supported value types are numbers,
enumerations, booleans, strings, raw bytes, or other nested message
types.  The \code{.proto} file syntax for defining the structure of Protocol
Buffer data is described comprehensively on Google Code\footnote{See 
\url{http://code.google.com/apis/protocolbuffers/docs/proto.html}.}.
Table~\ref{tab:proto} shows an example \code{.proto} file that
defines the \code{tutorial.Person} type\footnote{The compound name
  \code{tutorial.Person} in R is derived from the name of the
  message (\emph{Person}) and the name of the package defined at the top of the
  \code{.proto} file in which it is defined (\emph{tutorial}).}.  The \proglang{R} code in the right
column shows an example of creating a new message of this type and
populating its fields.

\noindent
\begin{table}
\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
\toprule
Schema : \code{addressbook.proto} & Example \proglang{R} session\\
\cmidrule{1-2}
\begin{minipage}{.40\textwidth}
\vspace{2mm}
\begin{example}
package tutorial;
message Person {
  required string name = 1;
  required int32 id = 2;
  optional string email = 3;
  enum PhoneType {
    MOBILE = 0; 
    HOME = 1;
    WORK = 2;
  }
  message PhoneNumber {
    required string number = 1;
    optional PhoneType type = 2;
  }
  repeated PhoneNumber phone = 4;
}
\end{example}
\vspace{2mm}
\end{minipage} & \begin{minipage}{.55\textwidth}
<<echo=TRUE>>=
library("RProtoBuf")
p <- new(tutorial.Person, id=1,
         name="Dirk")
p$name
p$name <- "Murray"
cat(as.character(p))
serialize(p, NULL)
class(p)
@
\end{minipage} \\
\bottomrule
\end{tabular}
\caption{The schema representation from a \code{.proto} file for the
  \code{tutorial.Person} class (left) and simple \proglang{R} code for creating
  an object of this class and accessing its fields (right).}
\label{tab:proto}
\end{table}

@@

% Note: We comment out subsections in favor of textbf blocks to save
% space and shrink down this section a little bit.
%\subsection[Importing message descriptors from .proto files]{Importing message descriptors from \code{.proto} files}

\subsection*{Importing message descriptors from \code{.proto} files}

>>>>

\documentclass[article]{jss}
\usepackage{thumbpdf,listings} 
\graphicspath{{Figures/}}
\shortcites{sciencecloud,janus,dremel,nlme}

\DefineVerbatimEnvironment{example}{Verbatim}{}

\author{Dirk Eddelbuettel\\Debian Project \And 
        Murray Stokely\\Google, Inc. \And
        Jeroen Ooms\\University of California,\\Los Angeles}
\Plainauthor{Dirk Eddelbuettel, Murray Stokely, Jeroen Ooms}

\title{\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in \proglang{R}}
\Plaintitle{RProtoBuf: Efficient Cross-Language Data Serialization in R}
\Shorttitle{\pkg{RProtoBuf}: Protocol Buffers in \proglang{R}}

\Abstract{
  Modern data collection and analysis pipelines often involve
  a sophisticated mix of applications written in general purpose and
  specialized programming languages.  Many formats commonly used to
  import and export data between different programs or systems, such
  as CSV or JSON, are verbose, inefficient, not
  type-safe, or tied to a specific programming language.  Protocol
  Buffers are a popular method of serializing structured data between
  applications -- while remaining independent of programming languages
  or operating systems.  They offer a unique combination of features,
  performance, and maturity that seems particularly well suited for
  data-driven applications and numerical computing.  The
  \pkg{RProtoBuf} package provides a complete interface to Protocol
  Buffers from the \proglang{R} environment for statistical computing.
  This paper outlines the general class of data serialization
  requirements for statistical computing, describes the implementation
  of the \pkg{RProtoBuf} package, and illustrates its use with example
  applications in large-scale data collection pipelines and web
  services.
}

\Keywords{\proglang{R}, \pkg{Rcpp}, Protocol Buffers, serialization, cross-platform}
\Plainkeywords{R, Rcpp, Protocol Buffers, serialization, cross-platform}

\Volume{71}
\Issue{2}
\Month{July}
\Year{2016}
\Submitdate{2014-02-05}
\Acceptdate{2015-04-13}
\DOI{10.18637/jss.v071.i02}

\Address{
  Dirk Eddelbuettel \\
  Debian Project \\
  River Forest, IL, United States of America\\
  E-mail: \email{edd@debian.org}\\
  URL: \url{http://dirk.eddelbuettel.com/}\\

  Murray Stokely\\
  Google, Inc.\\
  1600 Amphitheatre Parkway\\
  Mountain View, CA, United States of America\\
  E-mail: \email{mstokely@google.com}\\
  URL: \url{http://www.stokely.org/}\\

  Jeroen Ooms\\
  UCLA Department of Statistics\\
  University of California, Los Angeles\\
  Los Angeles, CA, United States of America\\
  E-mail: \email{jeroen.ooms@stat.ucla.edu}\\
  URL: \url{https://jeroenooms.github.io/}
}

\begin{document}

\vspace*{-0.3cm}

@@

Modern data collection and analysis pipelines increasingly involve collections
of decoupled components in order to better manage software complexity 
through reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
These pipelines are frequently built using different programming 
languages for the different phases of data analysis -- collection,
cleaning, modeling, analysis, post-processing, and
presentation -- in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may produce intermediate results that need to be
stored in a file, or sent over the network for further processing.

Given these requirements, how do we safely and efficiently share
intermediate results between different applications, possibly written
in different languages, and possibly running on different computer
systems?  In computer programming, \emph{serialization} is the process
of translating data structures, variables, and session states into a
format that can be stored or transmitted and then later reconstructed
in the original form \citep{clinec++}.  Programming languages such as
\proglang{R} \citep{r}, \proglang{Julia} \citep{julia},
\proglang{Java}, and \proglang{Python} \citep{python} include built-in
support for serialization, but the default formats are usually
language-specific and thereby lock the user into a single environment.

Data analysts and researchers often use character-separated text
formats such as CSV \citep{shafranovich2005common} to export
and import data. However, anyone who has ever used CSV files
will have noticed that this method has many limitations: It is
restricted to tabular data, lacks type-safety, and has limited
precision for numeric values.  Moreover, ambiguities in the format
itself frequently cause problems.  For example, conventions on which
characters are used as separator or decimal point vary by country.
\emph{Extensible markup language} (XML) is a well-established
and widely-supported format with the ability to define just about any
arbitrarily complex schema \citep{nolan2013xml}. However, it pays for
this complexity with comparatively large and verbose messages, and
added complexity at the parsing side (these problems are somewhat
mitigated by the availability of mature libraries and
parsers). Because XML is text-based and has no native notion of
numeric types or arrays, it is usually not a very practical format to
store numeric data sets as they appear in statistical applications.

A more modern format is \emph{\proglang{JavaScript} object notation}
(JSON), which is derived from the object literals of
\proglang{JavaScript}, and already widely-used on the world wide web.
Several \proglang{R} packages implement functions to parse and
generate JSON data from \proglang{R} objects
\citep{rjson,RJSONIO,jsonlite}.  JSON natively supports arrays
and four primitive types: numbers, strings, booleans, and
null. However, as it also is a text-based format, numbers are stored
in human-readable decimal notation which is inefficient and leads to
loss of type (double versus integer) and precision.  A number of
binary formats based on JSON have been proposed that reduce the
parsing cost and improve efficiency, but these formats are not widely
supported.  Furthermore, such formats lack a separate schema for the
serialized data and thus still duplicate field names with each message
sent over the network or stored in a file.

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift \citep{Apache:Thrift}, and Apache Avro \citep{Apache:Avro}
provide a compact well-documented schema for cross-language data
structures and efficient binary interchange formats.  Since the schema
is provided separately from the data, the data can be
efficiently encoded to minimize storage costs when
compared with simple ``schema-less'' binary interchange formats.
%Many sources compare data serialization formats
%and show Protocol Buffers perform favorably to the alternatives; see
%\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.
Protocol Buffers perform well in the comparison of such formats by
\citet{Sumaray:2012:CDS:2184751.2184810}.

This paper describes an \proglang{R} interface to Protocol Buffers,
and is organized as follows. Section~\ref{sec:protobuf}
provides a general high-level overview of Protocol Buffers as well as a basic
motivation for their use.
Section~\ref{sec:rprotobuf-basic} describes the interactive \proglang{R} interface
provided by the \pkg{RProtoBuf} package, and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
details the implementation of the main \proglang{S}4 classes and methods.
Section~\ref{sec:types} describes the challenges of type coercion
between \proglang{R} and other languages.  Section~\ref{sec:evaluation} introduces a
general \proglang{R} language schema for serializing arbitrary \proglang{R} objects and compares it to
the serialization capabilities built directly into \proglang{R}.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \pkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.

@@

Protocol Buffers are a modern, language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data. Some of their
features, particularly in the context of data analysis, are:
%
\begin{itemize}
\item \emph{Portable}:  Enable users to send and receive data between
  applications as well as different computers or operating systems.
\item \emph{Efficient}:  Data is serialized into a compact binary
  representation for transmission or storage.
\item \emph{Extensible}:  New fields can be added to Protocol Buffer schemas
  in a forward-compatible way that does not break older applications.
\item \emph{Stable}:  Protocol Buffers have been in wide use for over a
  decade.
\end{itemize}
%
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{protobuf-distributed-system-crop.pdf}
\caption{Example usage of Protocol Buffers.}
\label{fig:protobuf-distributed-usecase}
\end{figure}
%
Figure~\ref{fig:protobuf-distributed-usecase} illustrates an example
communication work flow with Protocol Buffers and an interactive \proglang{R} session.
Common use cases include populating a request remote-procedure call (RPC)
Protocol Buffer in \proglang{R} that is then serialized and sent over the network to a
remote server.  The server deserializes the message, acts on the
request, and responds with a new Protocol Buffer over the network.
The key difference to, say, a request to an \pkg{Rserve}
\citep{Urbanek:2003:Rserve,CRAN:Rserve} instance is that
the remote server may be implemented in any language.
%, with no dependence on \proglang{R}.

While traditional IDLs have at times been criticized for code bloat
and complexity, Protocol Buffers are based on a simple list and
records model that is flexible and easy to use.  The schema for
structured Protocol Buffer data is defined in \code{.proto} files,
which may contain one or more message types.  Each message type has
one or more fields.  A field is specified with a unique number (called
a \emph{tag number}), a name, a value type, and a field rule
specifying whether the field is optional, required, or repeated.  The
supported value types are numbers, enumerations, booleans, strings,
raw bytes, or other nested message types.  The \code{.proto} file
syntax for defining the structure of Protocol Buffer data is described
comprehensively on Google Code at
  \url{http://code.google.com/apis/protocolbuffers/docs/proto.html}.
Table~\ref{tab:proto} shows an example \code{.proto} file that defines
the \code{tutorial.Person} type\footnote{The compound name
  \code{tutorial.Person} in \proglang{R} is derived from the name of
  the message (\emph{Person}) and the name of the package defined at
  the top of the \code{.proto} file in which it is defined
  (\emph{tutorial}).}.  The \proglang{R} code in the right column
shows an example of creating a new message of this type and populating
its fields.

\begin{table}[t!]
\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
\hline
Schema : \code{addressbook.proto} & Example \proglang{R} session\\
\hline\\[-10pt]
\begin{minipage}{.40\textwidth}
\vspace{2mm}
\begin{example}
package tutorial;
message Person {
  required string name = 1;
  required int32 id = 2;
  optional string email = 3;
  enum PhoneType {
    MOBILE = 0; 
    HOME = 1;
    WORK = 2;
  }
  message PhoneNumber {
    required string number = 1;
    optional PhoneType type = 2;
  }
  repeated PhoneNumber phone = 4;
}
\end{example}
\vspace{2mm}
\end{minipage} & \begin{minipage}{.55\textwidth}
\begin{Schunk}
\begin{Sinput}
R> library("RProtoBuf")
R> p <- new(tutorial.Person, id = 1,
+    name = "Dirk")
R> p$name
\end{Sinput}
\begin{Soutput}
[1] "Dirk"
\end{Soutput}
\begin{Sinput}
R> p$name <- "Murray"
R> cat(as.character(p))
\end{Sinput}
\begin{Soutput}
name: "Murray"
id: 1
\end{Soutput}
\begin{Sinput}
R> serialize(p, NULL)
\end{Sinput}
\begin{Soutput}
 [1] 0a 06 4d 75 72 72 61 79 10 01
\end{Soutput}
\begin{Sinput}
R> class(p)
\end{Sinput}
\begin{Soutput}
[1] "Message"
attr(,"package")
[1] "RProtoBuf"
\end{Soutput}
\end{Schunk}
\vspace{0.5mm}
\end{minipage} \\
\hline
\end{tabular}
\caption{The schema representation from a \code{.proto} file for the
  \code{tutorial.Person} type (left) and simple \proglang{R} code for creating
  an object of this class and accessing its fields (right).}
\label{tab:proto}
\end{table}

@@

\subsection[Importing Message Descriptors from .proto files]{Importing
  Message Descriptors from \code{.proto} files}
